---
title: "Visualization of Crime in San Francisco"
author: "Kent Mok"
date: "February 4, 2016"
output: html_document
---

# Data Source
The data used in this analysis is from the crime dataset of Summer 2014 in San
Francisco. It can be found here: https://github.com/uwescience/datasci_course_materials/blob/master/assignment6/sanfrancisco_incidents_summer_2014.csv
Additional data for San Francisco can be found at: https://data.sfgov.org/

```{r load_packages, echo=FALSE, message=FALSE}
Installed <- require("ggplot2")
if (!Installed) {
    install.packages("ggplot2", repos = "http://cran.rstudio.com")
    library("ggplot2")
}

Installed <- require("ggmap")
if (!Installed) {
    install.packages("ggmap", repos = "http://cran.rstudio.com")
    library("ggmap")
}
```

```{r load_data, echo=FALSE}
rawdata <- read.csv("sanfrancisco_incidents_summer_2014.csv")
```

# Exploration of Data
A quick visualization of the data by summarizing the crime incidents by their
category shows that the "Larceny/Theft" category outnumbers the rest of the
categories by almost three times.
```{r, echo=FALSE}
crime_cat <- as.data.frame(table(rawdata$Category), stringsAsFactors = FALSE)
plotCategories <- ggplot(crime_cat)
plotCategories <- plotCategories +
    geom_bar(aes(Var1, Freq), stat = "identity", fill = "#0000CC") + 
    theme_bw() + 
    labs(x = "Crime", y = "Occurrences", title = "Frequency by Category") + 
    theme(axis.text.x=element_text(angle=90, size=9, hjust=1, vjust=0.5))
print(plotCategories)
```

```{r, echo=FALSE, eval=FALSE}
crime_day <- as.data.frame(table(rawdata$DayOfWeek), stringsAsFactors = FALSE)
plotDay <- ggplot(crime_day)
plotDay <- plotDay + 
    geom_bar(aes(Var1, Freq), stat = "identity", fill = "#0000CC") + 
    theme_bw() + 
    labs(x = "Day of Week", y = "Occurrences", title = "Frequency by Day") + 
    theme(axis.text.x=element_text(angle=90, size=9, hjust=1, vjust=0.5))
print(plotDay)
```

The top 10 occurrences of crimes by description are:
```{r, echo=FALSE, comment=NA}
crime_desc <- as.data.frame(table(rawdata$Descript), stringsAsFactors = FALSE)
names(crime_desc) <- c("Description", "Frequency")
print(crime_desc[order(-crime_desc$Frequency),][1:10,], row.names = FALSE)
```
Four of the top ten crimes by descriptions are categorized as "Larceny/Theft,"
so it makes sense that "Larceny/Theft" outnumbers all other crimes by so much.
The top crime by description is "Grand theft from locked auto." It occurred more
than three times the second top "Larceny/Theft" crime of "Petty theft of 
property." Since it accounts for the majority of "Larceny/Theft" crimes, it will
be the focus of the analysis.

```{r, echo=FALSE, eval=FALSE}
crime_desc[order(-crime_desc$Freq),][1,]
```

#Data Analysis
```{r, echo=FALSE, eval=FALSE}
gtadata <- rawdata[rawdata$Descript == "GRAND THEFT FROM LOCKED AUTO",]
crime_gtaday <- as.data.frame(table(gtadata$DayOfWeek),
                              stringsAsFactors = FALSE)
crime_gtaday$Var1 <- factor(crime_gtaday$Var1, c("Monday", "Tuesday", 
                                                 "Wednesday", "Thursday",
                                                 "Friday", "Saturday", 
                                                 "Sunday"))
plotGTADay <- ggplot(crime_gtaday)
plotGTADay <- plotGTADay + 
    geom_bar(aes(x = Var1, y = Freq), stat = "identity", fill = "#0000CC") + 
    theme_bw() + 
    labs(x = "Day of Week", y = "Occurrences", title = 
             "Grand Theft Auto by Day") + 
    theme(axis.text.x=element_text(angle=90, size=9, hjust=1, vjust=0.5))
print(plotGTADay)
```

A quick plot of the average occurrences of "Grand Theft From Locked Auto" by day
reveals that there is a slight increase in occurrences on Fridays, Saturdays, 
and Sundays over the other weekdays.
```{r, echo=FALSE}
gtadata <- rawdata[rawdata$Descript == "GRAND THEFT FROM LOCKED AUTO",]
crime_gtaday <- as.data.frame(table(gtadata$DayOfWeek),
                              stringsAsFactors = FALSE)
crime_gtaday$Var1 <- factor(crime_gtaday$Var1, c("Monday", "Tuesday", 
                                                 "Wednesday", "Thursday",
                                                 "Friday", "Saturday", 
                                                 "Sunday"))
crime_gtaday$Freq <- crime_gtaday$Freq / 92
plotGTADay <- ggplot(crime_gtaday)
plotGTADay <- plotGTADay + 
    geom_bar(aes(Var1, Freq), stat = "identity", fill = "#0000CC") + 
    theme_bw() + 
    labs(x = "Day of Week", y = "Average Occurrence Per Day", title = 
             "Grand Theft Auto by Day") + 
    theme(axis.text.x=element_text(angle=90, size=9, hjust=1, vjust=0.5))
print(plotGTADay)
```

This could possibly be attributed to more people going out in the city and
parking their vehicles in the city. If they leave any valuables in their 
vehicles, this would present more opportunities for break-ins and thefts. 

```{r, echo=FALSE, eval=FALSE}
crimeByTimeDay <- ggplot(gtadata)
crimeByTimeDay <- crimeByTimeDay +
    geom_point(aes(x = DayOfWeek, y = Time, alpha = 1/10), size = 0)
#, aes(x=sex, y=total_bill, fill=time)) +
#    geom_bar(stat="identity", position=position_dodge(), colour="black")
print(crimeByTimeDay)
```

A visualization of the crime occurrences by the hour of day shows a sharp 
increase during the hours of 6pm to 11pm.  
```{r, echo=FALSE}
crime_gtatime <- as.data.frame(table(gtadata$Time),
                              stringsAsFactors = FALSE)
crime_gtatime <- crime_gtatime[crime_gtatime$Freq != 0,]
#crime_gtatime$Hour <- as.numeric(substr(crime_gtatime$Var1, 1, 2))
crime_gtatime$Hour <- substr(crime_gtatime$Var1, 1, 2)
GTATimeOfDay <- ggplot(crime_gtatime, aes(x = Hour, Freq))
GTATimeOfDay <- GTATimeOfDay + 
    stat_summary(fun.y = sum,  # Adds up all observations for the timeframe
                 geom = "bar", fill = "#0000CC") +
    theme_bw() + 
    labs(x = "Time of Day", y = "Occurrences", title = 
             "Grand Theft Auto by Time of Day") + 
    theme(axis.text.x=element_text(angle=90, size=9, hjust=1, vjust=0.5)) +
    scale_x_discrete(labels = c(paste0("0",seq(0,9,1),":00"),paste0(seq(10,23,1),":","00")))
print(GTATimeOfDay)
```
  
This could possibly be attributed by the aforementioned increase of vehicles
with valuables parked in the city. Another contribution to the increase in
grand theft from locked vehicles could be more suitable conditions for the
activity due to sunset, which reduces available light and general visibility.
  
A plot of the locations of "Grand Theft From Locked Auto" incidents shows a high
concentration of incidents near the Tenderloin, Fisherman's Wharf, Embarcadero,
SOMA, and Cathedral Hill areas.
```{r, echo=FALSE,message=FALSE}
sfMap = map = get_map(location = 'San Francisco', zoom = 12)
gtadata$Bin <- substr(gtadata$Time,1,2)
gtadata$Bin <- cut(as.numeric(gtadata$Bin), breaks = c(0,6,12,18,24),
                    labels = c("00:00", "06:00", "12:00", "18:00"))
ggmap(sfMap) + geom_point(data = gtadata, aes(x = X, y = Y, color = Bin),
                          size = 0.5, alpha = 0.7) +
    scale_color_brewer(name = "Time of Day", 
                       labels = c("00:00-06:00", "06:00-12:00", "12:00-18:00",
                                  "18:00-24:00"), palette = "Set1") + 
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    labs(list(title = "Grand Theft Auto by Location and Time", x = "Longitude",
              y = "Latitude"))
```
  
These are all either tourist areas or areas with night life, which is in line
with the reasoning that the occurrence of incidents are related to the
opportunity from tourists visiting the city or others going to the city for a
night out and leaving valuables in the vehicle.

#Conclusion
This analysis shows that the highest crime incidents are thefts from locked
vehicles primarily in the hours between 6pm to 11pm. In order for the city to 
reduce this type of crime, it should inform tourists and visitors not to leave
any valuables in their vehicles when they are parked in the Tenderloin, 
Fisherman's Wharf, Embarcadero, SOMA, and Cathedral Hill areas. They would be 
well advised to be extra vigilant if they are parking and leaving their vehicles
between the hours of 6pm to 11pm.

#Reproducibility
The code used in this analysis can be found at:  
https://github.com/kmok23/DataScienceComm